### 2016-05-12 Draft.tex
Corrected 'partial finite-mixture'. Changed nomenclature for F_T(t|...). Changed T_{sq} to T_{sq}|det.

We distinguish between non-constant detection and detection heterogeneity, which can be modeled in similar ways but refer to different mechanisms.  
We reserve `detection heterogeneity' for differences across population subgroups.  
It is often modeled using a mixture of detection distributions.  
\citet{Farnsworth2002} introduce a partial finite-mixture for detection probabilities distinguishing between easy-to-detect individuals, which are always detected in first observation interval, and hard-to-detect individuals, which are detected at a constant rate.  
 Heterogeneity at the individual level can be modeled as random effects in the calculation of detection probabilities \citep{DorazioRoyle2003, Mantyniemi2005}.  
We use `non-constant detection' when detection rates for an organism change with time.  
Models constructed to address detection heterogeneity may provide reasonable estimates for non-constant detection scenarios and vice versa \citep{Mantyniemi2005}.

@@

In this study, we construct a time-to-detection N-mixture model for estimating abundance from removal-sampled point-count surveys.  
We explore model behavior through a series of simulations testing: (i) how well constant-detection models fit non-constant detection data and vice versa, (ii) the utility of a partial finite-mixtures to model detection heterogeneity involving a non-constant detection subgroup, and (iii) model performance with covariates and random effects affecting both abundance and detection.  
Finally, we apply our model to Ovenbird data from the Minnesota Forest Breeding Bird Project.

@@

\subsection{Data Model}
We assume that counts from different surveys are conditionally independent.  
We define $T_{sq}$ as a random variable for the first time at which an individual is detected.
We assume that detection times for individuals at any given survey are independent, identically distributed according to a TTDD with cdf $F_T(t)$ and pdf $f_T(t)$, which may depend on covariates.  
The probability of detection for an animal is thus $p_{s}^{(det)} = F_T(C_I)$ and the conditional distribution of observed detection times can be written $f_{T|det}(t|t_{sq}<C_I) = f_T(t) \textbf{I}(t<C_I) / F_T(C_I)$.  
Denoting $N_{s}$ as the total number animals present at a survey, the data model is:
\begin{align}\label{eq:datamodel}
n_{s}^{(obs)} &\sim \mbox{Binomial}\left(N_{s}, p_{s}^{(det)}\right)\\
\label{eq:Tsrq}T_{sq} &\sim f_{T|det}(t|t_{sq}<C_I)
\end{align}
When detection times are censored, we replace Eq. \ref{eq:Tsrq} with $\vn_{s} \sim \mbox{Multinomial}\left(n_{s}^{(obs)}, \textbf{p}_{s}\right)$, where $\textbf{p}_{s}$ is the vector of interval-specific detection probabilities obtained by partitioning the conditional cdf $F_{T|det}(t|t_{sq}<C_I)$ at the interval cutoff times.  
It follows that $\sum_{i=1}^I p_{si} = 1$.

@@

We can model detection heterogeneity by defining $F_T(t)$ as a mixture of TTDD components.  
In this study, we use interval-censored observations, so in addition to the four above TTDDs (with no mixture component), we consider partial finite-mixtures of a point mass during the first survey interval plus a continuous component from an exponential, gamma, Weibull, or lognormal distribution.  
We define a constant mixing parameter $\gamma \in [0,1]$ for the proportion of animals not distributed according to the point mass.

@@

Combining the distribution and expectation function for $T_{s}$, we can describe a partial-mixture TTDD cdf as:
\[F_T(t|\textbf{X}_{s}^D, \textbf{Z}_{s}^D, \boldsymbol{\theta}) = (1-\gamma)\mathbb{I}(t>0) + \gamma F_{T1}(t|\textbf{X}_{s}^D, \textbf{Z}_{s}^D, \boldsymbol{\theta})\]
where $F_{T1}$ is the cdf of the continuous component and $\boldsymbol{\theta}$ is a vector of parameters appropriate to $F_{T1}$.  
Note that parameters in a mixture TTDD do not share the exact same interpretation as equivalent parameters from a non-mixture TTDD, since they only pertain to the $\gamma$ proportion of individuals that follow $F_{T1}$.

@@

We conducted three simulation studies to explore the behavior of models with non-constant TTDDs.
In Sim1, we compared models with detection heterogeneity via a partial finite-mixture TTDD versus models without detection heterogeneity.
We simulated both mixture and non-mixture datasets from exponential, gamma, lognormal, and Weibull TTDD families.  
Because model behavior may differ between peaked and nonpeaked datasets, we simulated both for each TTDD family except exponential.  
Thus, combining mixtures, `peakedness' and TTDD families, we simulated datasets from 14 different TTDDs in total.  
In Sim2, we examined the impact of misspecifying the $F_{T1}$ family (e.g., fitting an exponential TTDD model to data generated from a Weibull TTDD).  
Of particular interest were scenarios with a time-varying $F_{T1}$ component in the data but a time-constant component in the inference model, and vice versa.
Based on findings from Sim1, we considered only mixture TTDD datasets, and we fit each with mixture TTDD models from all four families.  
In Sim3, we modeled a more complete scenario where both data generation and model fitting included covariates and random effects for abundance and detection.

@@

The Ovenbird analysis highlights the ambiguity that emerged from the simulation studies: the results were entirely consistent with data generated from a partial finite-mixture involving a constant-rate $F_{T1}$ TTDD, but they were also consistent with partial finite-mixtures involving a time-varying TTDD.
Simulation results demonstrate that the exponential mixture model fits exponential mixture data well, but it does not fit time-varying mixture data well.
At this point, we have no reliable tool for differentiating field data between the two TTDD types.

@@

The time-to-detection model as presented here in our study does not incorporate distance.  
Consequently, our definition of a TTDD necessarily represents an averaging across distance classes and a consequent loss of information if distance data are available.  
We can modify our model to account for distance.  
To be consistent with earlier distance-removal studies \citep{Farnsworth2005, Amundson2014}, we could: (i) define a time-to-\textit{availability} distribution $F_T^A(t)$ in the same manner that we have defined a TTDD in this study, (ii) modify Equation \ref{eq:pdet} to reflect the detection distance to individual $q$: $p_{sq}^{(det)} = g(d_q) F_T^A(C_I|\textbf{X}_{s}^D, \textbf{Z}_{s}^D, \boldsymbol{\theta})$, and (iii) tweak our data model accordingly.

However, our analysis makes possible a different strategy that is more consistent with a time-to-event conceptualization of the problem.  
Rather than define availability and perceptibility as independent detection events over the course of the observation period, we can define them at the event level.  
In terms of the model we have presented in this paper, this is equivalent to redefining the time-to-detection hazard function, $h_{sq}(t)$ --- which is now a function of the distance to individual $q$ --- as the product of the time-to-availability hazard function $h_{s}^A(t)$ and the perceptibility function $g(d/`8-_q)$.  
So, $h_{sq}(t) = g(d_q) h_{s}^A(t)$.  
From standard survival analysis results, $p_{sq}^{(det)} = 1 - \exp\left(-g(d_q)\int\limits_0^{C_I} h_{s}^A(u)du\right) = 1 - \left[1 - F_T^A(C_I|\textbf{X}_{s}^D, \textbf{Z}_{s}^D, \boldsymbol{\theta}) \right]^{g(d_q)}$.

>>>>

We distinguish between non-constant detection and detection heterogeneity, which can be modeled in similar ways but refer to different mechanisms.  
We reserve `detection heterogeneity' for differences across population subgroups.  
It is often modeled using a mixture of detection distributions.  
\citet{Farnsworth2002} introduce a finite-mixture for detection probabilities distinguishing between easy-to-detect individuals, which are always detected in first observation interval, and hard-to-detect individuals, which are detected at a constant rate.  
 Heterogeneity at the individual level can be modeled as random effects in the calculation of detection probabilities \citep{DorazioRoyle2003, Mantyniemi2005}.  
We use `non-constant detection' when detection rates for an organism change with time.  
Models constructed to address detection heterogeneity may provide reasonable estimates for non-constant detection scenarios and vice versa \citep{Mantyniemi2005}.

@@

In this study, we construct a time-to-detection N-mixture model for estimating abundance from removal-sampled point-count surveys.  
We explore model behavior through a series of simulations testing: (i) how well constant-detection models fit non-constant detection data and vice versa, (ii) the utility of finite-mixtures to model detection heterogeneity involving a non-constant detection subgroup, and (iii) model performance with covariates and random effects affecting both abundance and detection.  
Finally, we apply our model to Ovenbird data from the Minnesota Forest Breeding Bird Project.

@@

\subsection{Data Model}
We assume that counts from different surveys are conditionally independent.  
We define $T_{sq}$ as a random variable for the first time at which an individual is detected.
We assume that detection times for individuals at any given survey are independent, identically distributed according to a TTDD with cdf $F_T(t)$ and pdf $f_T(t)$, which may depend on covariates.  
The probability of detection for an animal is thus $p_{s}^{(det)} = F_T(C_I)$ and the conditional distribution of observed detection times can be written $f_{T|det}(t|t_{sq}<C_I) = f_T(t) \textbf{I}(t<C_I) / F_T(C_I)$.  
Denoting $N_{s}$ as the total number animals present at a survey, the data model is:
\begin{align}\label{eq:datamodel}
n_{s}^{(obs)} &\sim \mbox{Binomial}\left(N_{s}, p_{s}^{(det)}\right)\\
\label{eq:Tsrq}T_{sq}|det &\sim f_{T|det}(t|t_{sq}<C_I)
\end{align}
When detection times are censored, we replace Eq. \ref{eq:Tsrq} with $\vn_{s} \sim \mbox{Multinomial}\left(n_{s}^{(obs)}, \textbf{p}_{s}\right)$, where $\textbf{p}_{s}$ is the vector of interval-specific detection probabilities obtained by partitioning the conditional cdf $F_{T|det}(t|t_{sq}<C_I)$ at the interval cutoff times.  
It follows that $\sum_{i=1}^I p_{si} = 1$.

@@

We can model detection heterogeneity by defining $F_T(t)$ as a mixture of TTDD components.  
In this study, we use interval-censored observations, so in addition to the four above TTDDs (with no mixture component), we consider finite-mixtures of a point mass during the first survey interval plus a continuous component from an exponential, gamma, Weibull, or lognormal distribution.

@@

Combining the distribution and expectation function for $T_{s}$, we can describe a mixture TTDD cdf as:
\[F_T(t|\varphi_s, \boldsymbol{\theta}, \gamma) = (1-\gamma)\mathbb{I}(t>0) + \gamma F_{T1}(t|\varphi_s, \boldsymbol{\theta})\]
where $F_{T1}$ is the cdf of the continuous component, $\boldsymbol{\theta}$ is a vector of parameters appropriate to $F_{T1}$, and $\gamma \in [0,1]$ is the proportion of animals not distributed according to the point mass.
Note that parameters in a mixture TTDD do not share the exact same interpretation as equivalent parameters from a non-mixture TTDD, since they only pertain to the $\gamma$ proportion of individuals that follow $F_{T1}$.

@@

We conducted three simulation studies to explore the behavior of models with non-constant TTDDs.
In Sim1, we compared models with detection heterogeneity via a finite-mixture TTDD versus models without detection heterogeneity.
We simulated both mixture and non-mixture datasets from exponential, gamma, lognormal, and Weibull TTDD families.  
Because model behavior may differ between peaked and nonpeaked datasets, we simulated both for each TTDD family except exponential.  
Thus, combining mixtures, `peakedness' and TTDD families, we simulated datasets from 14 different TTDDs in total.  
In Sim2, we examined the impact of misspecifying the $F_{T1}$ family (e.g., fitting an exponential TTDD model to data generated from a Weibull TTDD).  
Of particular interest were scenarios with a time-varying $F_{T1}$ component in the data but a time-constant component in the inference model, and vice versa.
Based on findings from Sim1, we considered only mixture TTDD datasets, and we fit each with mixture TTDD models from all four families.  
In Sim3, we modeled a more complete scenario where both data generation and model fitting included covariates and random effects for abundance and detection.

@@

The Ovenbird analysis highlights the ambiguity that emerged from the simulation studies: the results were entirely consistent with data generated from a finite-mixture involving a constant-rate $F_{T1}$ TTDD, but they were also consistent with finite-mixtures involving a time-varying TTDD.
Simulation results demonstrate that the exponential mixture model fits exponential mixture data well, but it does not fit time-varying mixture data well.
At this point, we have no reliable tool for differentiating field data between the two TTDD types.

@@

The time-to-detection model as presented here in our study does not incorporate distance.  
Consequently, our definition of a TTDD necessarily represents an averaging across distance classes and a consequent loss of information if distance data are available.  
We can modify our model to account for distance.  
To be consistent with earlier distance-removal studies \citep{Farnsworth2005, Amundson2014}, we could: (i) define a time-to-\textit{availability} distribution $F_T^A(t)$ in the same manner that we have defined a TTDD in this study, (ii) modify the equation for detection probability to reflect the detection distance to individual $q$: $p_{sq}^{(det)} = g(d_q) F_T^A(C_I|\varphi_s, \boldsymbol{\theta})$, and (iii) tweak our data model accordingly.

However, our analysis makes possible a different strategy that is more consistent with a time-to-event conceptualization of the problem.  
Rather than define availability and perceptibility as independent detection events over the course of the observation period, we can define them at the event level.  
In terms of the model we have presented in this paper, this is equivalent to redefining the time-to-detection hazard function, $h_{sq}(t)$ --- which is now a function of the distance to individual $q$ --- as the product of the time-to-availability hazard function $h_{s}^A(t)$ and the perceptibility function $g(d_q)$.  
So, $h_{sq}(t) = g(d_q) h_{s}^A(t)$.  
From standard survival analysis results, $p_{sq}^{(det)} = 1 - \exp\left(-g(d_q)\int\limits_0^{C_I} h_{s}^A(u)du\right) = 1 - \left[1 - F_T^A(C_I|\varphi_s, \boldsymbol{\theta}) \right]^{g(d_q)}$.
